Designing Types for R, Empirically

### 2020-05-16 paper/main.tex
stuff

Our community builds, improves, and reasons about programming languages. To make
design decisions that benefit users, we need to understand the target language
as well as the real-world needs it answers. Often, we can appeal to our
intuition, as many languages are intended for general purpose programming tasks.
Unfortunately, intuition may fail when looking at domain-specific languages as
these languages were designed for a specific group of users to solve very
specific needs. This is the case of the data science language R.

@@

Retrofitting a type system to the R programming language would increase our
assurance in the result of data analysis. But, we are faced with two
challenges. First, it is unclear what would be the \emph{right} type system
for a language as baroque as R. For example, one of the most popular data
type, the \code{data.frame}, is manipulated through reflective operations---a data frame is a table whose columns can be added or removed on the fly.
Second, but just as crucially, designing a type system that will be adopted
would require overcoming some prejudices and educating large numbers of
users.

This paper is a data-driven study of what a type system for the R language
could look like. Our intention is to eventually propose changes to the
language, but we are aware that for any changes to be accepted by the user
community they must clearly benefit the language without endangering
backwards compatibility. Our goal is thus to find a compromise between
simplicity and usefulness; any proposed type system should cover common
programming idioms while remaining easy to learn and to use.

This paper focuses on a simpler problem than an entire type system, instead,
we limit the scope of our investigation to giving types to function
signatures. In order to do this, we design a simple type language, one that
matches the R data types but omits features such as parametric polymorphism
and subtyping between user-defined data types. We then extract type
signature from execution traces of a corpus of widely used libraries.  This
allows us to see how far one can get with the simple type language and
identify limitations of our this particular design. We validate the
robustness of the extracted type signature by implementing a contact system
that weaves the types around their respective functions, and use a large
number of clients of the target packages for validation. The contract system
can be used by the R community to experiment with our type signatures and to
replace ad hoc error checking code.

To sum up, our paper makes the following main contributions:
\begin{itemize}
\item We implemented tooling to automatically extract type signatures from R
  functions and to instrument R functions with checks based on their declared
  types. The tooling is robust and scalable to the entire R language.
\item We carried out a large-scale analysis of corpus of \PACKAGES widely
  used and actively maintained libraries to extract function type signatures and
  validated the robustness of the inferred type signatures against \NUMFILESRUN
  programs that use those functions.
\item We report on the appropriateness and usefulness of a simple type
  language for the R programming language.
\end{itemize}

@@

Recent work has gone into adding types to Python, the other eminent data science language.
Typilus is an interesting piece of recent work which explores using neural networks to infer types for Python programs~\cite{allamanis2020typilus}, and Python itself added support for type hints in Python 3.5~\cite{pythonspectypes}.
There is no previous work on types for the R programming language. We take
inspiration in the above mentioned works but focus on adapting them to our
target language.

@@

While this task is seemingly simple, the details and their proverbial devil
are surprisingly tricky to get right and to scale to long running programs.
To avoid starting from scratch, our implementation reuses an open source
dynamic analysis framework for R named \rdt~\cite{oopsla19} which consists
of an instrumented R Virtual Machine based on GNU-R version 3.5.0. The
framework exposes hooks at key points in the R interpreter to which user
defined callbacks can be attached. These hooks include function entry and
exit, method dispatch for the S3 and S4 object systems, the longjumps used
by the interpreter to implement non-local exit, creation and forcing of
promises (lazy evaluation), variable definition, value creation, mutation
and garbage collection.

\paragraph{Raw types}
The type information output by the tool includes the \emph{type tag} of each
value. The R internal types are translated to names in the proposed type
system. The next bit of information output is the \emph{class} which is an
optional list of names, it may be absent, and, in some cases, it may be
implicit (i.e. the interpreter blesses some values with the \k{matrix} and
\k{array} classes even when they have no attributes). 
Depending on the value's type tag, the tool collects further information:

\begin{itemize}
\item For vectors, the number of dimensions and the present of \NA values.
\item For lists, a recursive traversal collects element types.
\item For promises, we attempt to establish the type of its result or output
  \any.
\end{itemize}

\noindent
To obtain raw types, we make use of R's C FFI and use low-level machinery to
collect type tags and attributes from the R run-time.  The types that we
provide to users are constructed during post-processing, and rely on the
detailed information made available by these low-level reflection
mechanisms.

\paragraph{Varargs}
Arguments that are part of a function's varargs (denoted \VARGS) are
ignored, as type \VARGS is output as the type of the entire varargs construct.

\paragraph{Promises}
The fact that arguments are evaluated lazily (in that expression are packed
into promises and evaluated on first access) complicates the information
gathering.  For example, some promises remain un-evaluated, and it would be
erroneous to force them as they may cause side-effects that will affect the
programs execution.  To deal with unevaluated arguments, we make an initial
guess for each argument at function entry.  If the promise is later forced,
we simply update the recorded type for the argument.

@@

\paragraph{Non-local returns}
When a function exits with a longjump there is no return value to speak
of. In order to ensure that call traces are valid when a longjump occurs, we
intercept the unwinding process initiated by the longjump and mimic the
functionality of the functions being exited. When no return value is present
for a call, we record a special \k{jumped} return type.

@@

\noindent
In other words, we take the union of the types occurring at individual
argument positions rather than an union of function types. Furthermore, we
apply some transformation on the types to keep the size of types in check.

\medskip
\begin{tabular}{rclll}
  \T \k{~|~} \T  & $\Rightarrow$ &  \T &  \\
  \T \k{~|~} $T'$ & $\Rightarrow$ &  \T & \it{iff} $T' <: T$  \\
  \LIST{\T} \k{~|~} $\LIST{T'}$ & $\Rightarrow$ &  \LIST\T & \it{iff} $T' <: T$  \\
   $\null \k{~|~} \VEC{\S_1} \k{~|~}\dots \VEC{\S_n}$  & $\Rightarrow$ &
  $ \NAVEC{\S_1} \k{~|~}\dots \NAVEC{\S_n}$ &  \\
  \\
  \\
  \NAVEC\S  & <:& \VEC\S \\
  \T        & <:& \k{?}\T \\
  \LIST{\T} & <:& \LIST{\T'} & \it{iff} \T <: \T' \\
  \S        & <:& \VEC\S \\
  \lgl      & <:& \int\\
  \int      & <:& \dbl\\
  \dbl      & <:& \clx\\  
\end{tabular}
\medskip

\noindent
Assuming that type sequences can be reordered freely, we rewrite types to
minimize their size by removing redundant types, types that are subsumed by
subtyping, and \null types. For performance reasons, we bound lists to no
more than five types in their union, above that they become \LIST{any}.
Functions are alos limited to twenty arguments. Higher-order functions are
approximated by \any to \any.

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Evaluation} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand{\FIX}[1]{{\bf #1}\xspace}

We ran \typetracer on the test, example, and vignette code of the aforementioned corpus of \PACKAGES packages and inferred types for \NUMFUNCTIONS functions.
We have selected and will discuss a few interesting signatures.

\begin{table}[H]
  \centering
  \resizebox{.8\linewidth}{!}{
  \begin{tabular}{llrrr}
    \toprule
    Function & Type Signature \\
    \midrule
    \code{dplyr::group_indices} & \FUN{\CLASS{\texttt{data.frame}}, \VARGS}{\VEC\int}\\
    \code{moments::all.cumulants} & \FUN{\CLASS{\texttt{matrix}} \; | \; \VEC\dbl}{\CLASS{\texttt{matrix}}\; | \; \VEC{\dbl}}\\
    \code{diptest::dip}& \FUN{\VEC\dbl, \chr \; | \; \lgl, \lgl,
                         \dbl}{\CLASS{\texttt{dip}} \; | \; \dbl} \\
    \code{stabledist::cospi2}& \FUN{\VEC\dbl}{\VEC\dbl} \\
    \code{matrixcalc::matrix.power}& \FUN{\CLASS{\texttt{matrix}},
                                     \dbl}{\CLASS{\texttt{matrix}}} \\
    \code{data.tree::Traverse}& \FUN{\CLASS{\texttt{Node}, \texttt{R6}}, ~\VEC\chr,
                                \any, \any}{\LIST{\any}} \\
    \code{openssl::decrypt\_envelope}& \FUN{~\VEC\raw, ~\VEC\raw, ~\VEC\raw,
                                       \CLASS{\texttt{key}, \texttt{rsa}}, \any}{\VEC\raw} \\
    \code{dbplyr::set\_win\_current\_group}& \FUN{? ~\VEC\chr}{\; ? \; \VEC\chr} \\
    \code{openssl::sha256}& \FUN{~\VEC\raw, ? ~\VEC\raw}{\VEC\raw} \\
    \code{forecast::initparam}& \FUN{? \dbl, \any, \any, \any, \chr, \chr, \lgl,
                                ~\VEC\dbl, ~\VEC\dbl, \any}{\VEC\dbl} \\
    \bottomrule
  \end{tabular}}
\caption{Select Type Signatures}\label{tbl:typesig}
\end{table}

Many of the features of our type language are represented here.
Some of these signatures are telling of the function's behaviour.
For example, consider the signature for \code{openssl::decrypt_envelope}: the first three parameters of the function are byte arrays, and the fourth argument is an RSA key, used to decrypt some of the inputs, and the output of the function is the decrypted byte array.
As another example, consider \code{data.tree::Traverse}: according to the function documentation, it takes the root of a tree and traverses it in an order specified by the second argument.
We see that reflected in the type, where the first argument has type \CLASS{\texttt{Node}, \texttt{R6}} and the second argument had type \VEC\chr, representing the traversal order.

>>>>

Our community builds, improves, and reasons about programming languages. To
make design decisions that benefit users, we need to understand our target
language as well as the real-world needs it answers. Often, we can appeal to
our intuition, as many languages are intended for general purpose
programming tasks.  Unfortunately, intuition may fail when looking at
domain-specific languages as these languages were designed for a specific
group of users to solve very specific needs. This is the case of the data
science language R.

@@

Retrofitting a type system to the R language would increase our assurance in
the result of data analysis. But, we are faced with two challenges. First,
it is unclear what would be the \emph{right} type system for a language as
baroque as R. For example, one of the most popular data type, the
\code{data.frame}, is manipulated through reflective operations---a data
frame is a table whose columns can be added or removed on the fly.  Second,
but just as crucially, designing a type system that will be adopted would
require overcoming some prejudices and educating large numbers of users.

This paper is a data-driven study of what a type system for the R language
could look like. Our intention is to eventually propose changes to the
language, but we are aware that for any changes to be accepted by the user
community they must clearly benefit the language without endangering
backwards compatibility. Our goal is thus to find a compromise between
simplicity and usefulness; any proposal should cover common programming
idioms while remaining easy to learn and to use.

This paper focuses on a simpler problem than an entire type system, instead,
we limit our scope to giving types to function signatures. For this, we
design a simple type language, one that matches the R data types but omits
features such as parametric polymorphism and subtyping between user-defined
data types. We then extract type signature from execution traces of a corpus
of widely used libraries.  This allows us to see how far one can get with
the simple type language and identify limitations of the design. We validate
the robustness of the extracted type signature by implementing a contact
checker that weaves the types around their respective functions, and use a
large number of clients of the target packages for validation. The contract
system can be used by the R community to experiment with type signatures and
replace ad hoc error checking code.

To sum up, our paper makes the following contributions:
\begin{itemize}
\item We implemented tooling to automatically extract type signatures from R
  functions and to instrument R functions with checks based on their declared
  types. The tooling is robust and scalable to the entire R language.
\item We carried out a large-scale analysis of corpus of \PACKAGES widely
  used and actively maintained libraries to extract function type signatures and
  validated the robustness of the inferred type signatures against \NUMFILESRUN
  programs that use those functions.
\item We report on the appropriateness and usefulness of a simple type
  language for the R programming language.
\end{itemize}

@@

Recent work has gone into adding types to Python, the other eminent data
science language.  Typilus is an interesting piece of recent work which
explores using neural networks to infer types for Python
programs~\cite{allamanis2020typilus}, and Python itself added support for
type hints in Python 3.5~\cite{pythonspectypes}.  There is no previous work
on types for R. We take inspiration in the above mentioned works but focus
on adapting them to our target language.

@@

While this task is seemingly simple, the details and their proverbial devil
are surprisingly tricky to get right and to scale to large, long running,
programs.  To avoid starting from scratch, our implementation reuses \rdt,
an open source dynamic analysis framework for R~\cite{oopsla19} which
consists of an instrumented R Virtual Machine based on GNU-R version
3.5.0. The framework exposes hooks in the interpreter to which user defined
callbacks can be attached. These hooks include function entry and exit,
method dispatch for the S3 and S4 object systems, the longjumps used by the
interpreter to implement non-local exit, creation and forcing of promises,
variable definition, value creation, mutation and garbage collection.

\paragraph{Raw types}
The type information output by the tool includes the \emph{type tag} of each
value. Internal types are translated to names in the proposed type
system. The next bit of informatio is the \emph{class}, an optional list of
names that may be absent, and, in some cases, is implicit (i.e. the
interpreter blesses some values with the \k{matrix} and \k{array} classes
even without attributes).  Depending on a value's type, the tool collects
further information:

\begin{itemize}
\item For vectors, the presence of \NA values.
\item For lists, element types by a recursive traversal.
\item For promises, an approximation of the expression type.
\end{itemize}

\noindent
To obtain these raw types, we make use of R's C FFI and use low-level
machinery to collect information from the R language run-time.  Types are
conmpleted during post-processing, and rely on the detailed information made
available by these low-level reflection mechanisms.

\paragraph{Promises}
The fact that arguments are lazy (expression are packed into promises and
only evaluated on first access) complicates information gathering.  For
example, some promises may remain un-evaluated, and it would be erroneous to
force them as they may side-effect and change programs execution.  To deal
with unevaluated arguments, we make an initial guess for each argument at
function entry.  If the promise is later forced, we simply update the
recorded type for the argument.

@@

\paragraph{Non-local returns}
When a function exits with a longjump, there is no return value to speak
of. To ensure call traces are valid when a longjump occurs, we intercept the
unwinding process and record a special \k{jumped} return type for function
returns that are skipped.

\paragraph{Varargs}
Arguments that are part of a function's varargs (denoted \VARGS) are
ignored. We do not attempt to give varags a type in this type system.

@@

\noindent
In other words, we take the union of the types occurring at individual
argument positions rather than an union of function types. Furthermore, we
apply some transformation on the types to keep the size of types in check.
The following overviews the main simplification rules as well as the simple
notion of subtyping that we have adopted here.

\medskip
\begin{tabular}{rclll}
  \T \k{~|~} \T  & $\Rightarrow$ &  \T &  \\
  \T \k{~|~} $T'$ & $\Rightarrow$ &  \T & \it{iff} $T' <: T$  \\
  \LIST{\T} \k{~|~} $\LIST{T'}$ & $\Rightarrow$ &  \LIST\T & \it{iff} $T' <: T$  \\
   $\null \k{~|~} \VEC{\S_1} \k{~|~}\dots \VEC{\S_n}$  & $\Rightarrow$ &
  $ \NAVEC{\S_1} \k{~|~}\dots \NAVEC{\S_n}$ &  \\
  \\
  \\
  \\
  \NAVEC\S  & <:& \VEC\S \\
  \T        & <:& \k{?}\T \\
  \LIST{\T} & <:& \LIST{\T'} & \it{iff} \T <: \T' \\
  \S        & <:& \VEC\S \\
  \lgl      & <:& \int\\
  \int      & <:& \dbl\\
  \dbl      & <:& \clx\\  
\end{tabular}
\medskip

\noindent
Assuming that type sequences can be reordered freely, we rewrite types to
minimize their size by removing redundant types, types that are subsumed by
subtyping, immutable lists, and remove \null type to replace them with
nullables. For performance reasons, we bound lists to no more than five
types in their union, above that they become \LIST{any}.  Functions are alos
limited to twenty arguments. Higher-order functions are approximated by \any
to \any.

@@

\newpage

@@

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Evaluation} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We ran \typetracer on the test, example, and vignette code of the
aforementioned corpus of \PACKAGES packages and successfully inferred types
for \NUMFUNCTIONS functions.  Table~\ref{tbl:typesig} illustrates the
process with ten representative signatures.  Many of the features of our
type language are represented here.  Some signatures are telling of the
function's behaviour.  For example, consider \code{decrypt_envelope}: the
firs three parameters of the function are byte arrays, and the fourth
argument is an RSA key, used to decrypt some of the inputs, and the output
of the function is another byte array.  As another example, consider
\code{Traverse}: according to the function documentation, it takes the root
of a tree and traverses it in an order specified by the second argument.  We
see that reflected in the type, where the first argument has type
\CLASS{\texttt{Node}, \texttt{R6}} and the second argument had type
\VEC\chr, representing the traversal order.

\begin{table}[H]
  \centering
  \resizebox{.98\linewidth}{!}{
  \begin{tabular}{llrrr}
    \toprule
    Function & Type Signature \\
    \midrule
    \code{dplyr::group_indices} & \FUN{\CLASS{\texttt{data.frame}}, \VARGS}{\VEC\int}\\
    \code{moments::all.cumulants} & \FUN{\CLASS{\texttt{matrix}} \; | \; \VEC\dbl}{\CLASS{\texttt{matrix}}\; | \; \VEC{\dbl}}\\
    \code{diptest::dip}& \FUN{\VEC\dbl, \chr \; | \; \lgl, \lgl,
                         \dbl}{\CLASS{\texttt{dip}} \; | \; \dbl} \\
    \code{stabledist::cospi2}& \FUN{\VEC\dbl}{\VEC\dbl} \\
    \code{matrixcalc::matrix.power}& \FUN{\CLASS{\texttt{matrix}},
                                     \dbl}{\CLASS{\texttt{matrix}}} \\
    \code{data.tree::Traverse}& \FUN{\CLASS{\texttt{Node}, \texttt{R6}}, ~\VEC\chr,
                                \any, \any}{\LIST{\any}} \\
    \code{openssl::decrypt\_envelope}& \FUN{~\VEC\raw, ~\VEC\raw, ~\VEC\raw,
                                       \CLASS{\texttt{key}, \texttt{rsa}}, \any}{\VEC\raw} \\
    \code{dbplyr::set\_win\_current\_group}& \FUN{? ~\VEC\chr}{\; ? \; \VEC\chr} \\
    \code{openssl::sha256}& \FUN{~\VEC\raw, ? ~\VEC\raw}{\VEC\raw} \\
    \code{forecast::initparam}& \FUN{? \dbl, \any, \any, \any, \chr, \chr, \lgl,
                                ~\VEC\dbl, ~\VEC\dbl, \any}{\VEC\dbl} \\
    \bottomrule
  \end{tabular}}
\caption{Select Type Signatures}\label{tbl:typesig}
\end{table}

This section attempts to evaluate how well the proposed type system is able
to describe the actual type signatures of functions. For this we focus on
how often there is a single type for a particular argument; this because
union types and \any are less accurate (and would likely require a more
refined notion of subtyping or parametric polymorphism). Then, we evaluate
how robust the inferred signatures are by checking that they remain valid
for other inputs.  Lastly, we try to see if the current proposal would be
useful to programmers by allowing them to remove ad hoc checks and providing
useful documentation.
